---
title: "quickr semantics: types, shapes, and differences from R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{quickr semantics: types, shapes, and differences from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

quickr compiles a subset of R. This vignette documents the semantics of
that subset: what type a result has, when shapes are checked, and every
place where a compiled function deliberately behaves differently from R.
Most deliberate boundaries take the form of an error where R would return
a value. The known value, type, and evaluation differences are documented
separately below.

## The contract

> Except for the documented non-error differences below,
> `quick(f)(x)` either returns exactly what `f(x)` returns — same values,
> same `typeof()`, same shape — or raises an error. Never a third thing.
> The unchecked `NA` inputs and out-of-bounds runtime subscripts documented
> below are outside this contract and have undefined behavior.

When quickr cannot reproduce R's behavior, it refuses. It prefers to
refuse at compile time, when `quick()` is called; when the decision
depends on values only known at run time (for example, sizes declared as
`NA` or as symbolic constraints like `double(n)`), the compiled function
performs a cheap runtime check and raises an error on mismatch. The text
of these error messages differs from R's.

The rest of this vignette is that contract in detail: first result
*types*, then operand *shapes*, then the list of deliberate divergences.

## Types

Value types follow R's promotion order for the supported atomic modes:

```
logical < integer < double < complex
```

Character values are not supported (declaring a `character` argument is
a compile error). Complex values support elementwise arithmetic and
equality comparisons; order comparisons (`<`, `<=`, `>`, `>=`) and `%%`
are errors, exactly as they are in R. Outside elementwise operations,
complex support is narrower than the lattice suggests — see
[complex values](#complex) under the differences below.
Raw values cannot be used as arithmetic operands.

An operation's result type is the "join" of its operands' types — the
highest type among them — except where the table says otherwise:

| Operation | Result type |
|---|---|
| `+` `-` `*` (and unary `+` `-`) | join of the operand types; logicals count as integers (in R, `TRUE + TRUE` is `2L`) |
| `/` | double (complex when an operand is complex, as in R) |
| `^` | double (complex when an operand is complex, as in R; see note below) |
| `%%`, `%/%` | join of the operand types; logicals count as integers |
| `<` `<=` `>` `>=` `==` `!=` | logical |
| `&`, `|`, `!`, `&&`, `||` | logical (see [logical operators](#logical-operators) below) |
| `c()` | join of all argument types (`c(TRUE, FALSE)` stays logical, as in R) |
| multi-argument `min()` / `max()` / `sum()` | join of all argument types; logicals count as integers (in R, `sum(TRUE, TRUE)` is `2L`; see [integer `sum()` overflow](#integer-sum-overflow)) |
| `prod()` | join of the argument types; unlike R, an integer or logical product is integer (see [`prod()`](#prod)) |
| single-argument `min()`, `max()`, and `sum()` | type of `x`; logical counts as integer |
| `abs()` | type of `x`, except complex input returns double; logical counts as integer |
| `ifelse(test, yes, no)` | join of `yes` and `no` (but see [ifelse](#ifelse) below) |
| `%*%`, `crossprod()`, `solve()`, and other linear algebra | always double, as in R; complex operands are a compile error (see [complex values](#complex)) |
| `t(x)`, `diag(x)` | same type as `x`; the identity forms `diag(n)` are double, as in R (see [scalar `t()`](#scalar-transpose)) |
| `cbind()` / `rbind()` | join of all argument types |

For example, mixing integer and double promotes:

```{r}
f <- quick(function(x, y) {
  declare(type(x = integer(1)), type(y = integer(1)))
  x / y
})
f(5L, 2L)
#> [1] 2.5
```

and logical arithmetic produces integers, as in R:

```{r}
f <- quick(function(x, y) {
  declare(type(x = logical(1)), type(y = logical(1)))
  x + y
})
r <- f(TRUE, TRUE)
r
#> [1] 2
typeof(r)
#> [1] "integer"
```

**A note on `^`:** the result is always double, as in R. When the
exponent has integer type, quickr computes the power with an integer
exponent. This is exact and well-defined for negative bases:
`(-2) ^ 3L` is exactly `-8`. A whole-valued *double* exponent remains a
real exponent; see [real exponents](#real-exponents) below.

### Reassignment cannot narrow a type

R happily re-binds a name to a value of a wider type: after
`x <- x + 0.5`, an integer `x` becomes double. Compiled code cannot
re-type a variable, and silently truncating the value would change
results, so quickr makes this a compile-time error:

```{r}
quick(function(x) {
  declare(type(x = integer(1)))
  x <- x + 0.5
  x
})
#> Error: cannot reassign `x`: assignment would narrow double to integer;
#>   R would promote `x` to double
```

The same check applies to subassignment (`x[i] <- ...`) and
superassignment (`x <<- ...`, `x[i] <<- ...`). Widening in the other
direction (assigning an integer value to a double variable) is fine and
inserts the conversion R would perform.

## Shapes

Every operation that combines two or more values checks that their
shapes are compatible. A shape question has one of three answers, and
each answer has a fixed consequence:

- **Provably compatible** at compile time — allowed, no check emitted.
- **Provably incompatible** at compile time — compile error.
- **Unknown** at compile time (sizes declared `NA`, or symbolic sizes
  that cannot be proven equal) — the compiled function checks at run
  time and raises an error on mismatch.

There is no case where a mismatch is silently accepted.

For elementwise operations (arithmetic, comparisons, `&`, `|`),
"compatible" means:

| Operands | Behavior |
|---|---|
| scalar with a nonempty array | allowed; the scalar is broadcast |
| identical shapes | allowed |
| vector with matrix, `length(vec) == nrow(mat)` | allowed; the vector is recycled column-wise, exactly as in R |
| a `1x1` matrix in *arithmetic* with a vector of statically known length other than 1 | recycled; the result is a plain vector without dims, exactly as in R (which warns that this recycling is deprecated) |
| a `1x1` matrix in comparisons and `&` / `|` | treated as a one-row matrix; mismatched shapes are rejected, and R errors here too |
| anything else — unequal lengths, even when one divides the other | **error**: R-style recycling is not supported |

A `1x1` matrix meeting a vector whose length is only known at run time
is the one place where R's *result shape* depends on a runtime value: R
keeps the `1x1` dims when the vector has length 1 and drops them for
any other length. A compiled function's result shape cannot depend on a
value, so quickr treats the `1x1` as a one-row matrix there instead:
the compiled function checks that the vector has length 1 and returns a
`1x1` matrix — matching R exactly — and errors on a longer vector,
where R would perform the deprecated recycling.

The last row is the important one. In R, `1:6 + c(10, 20)` recycles the
shorter vector. quickr never does this:

```{r}
quick(function(x, y) {
  declare(type(x = double(4)), type(y = double(2)))
  x + y
})
#> Error: elementwise vector operations require equal lengths or a scalar
#>   operand; R-style recycling is not supported
```

When lengths are not known at compile time, the same rule is enforced at
run time:

```{r}
f <- quick(function(x, y) {
  declare(type(x = double(n)), type(y = double(m)))
  x + y
})
f(c(1, 2), c(3, 4))
#> [1] 4 6
f(c(1, 2, 3), c(1, 2))
#> Error: elementwise vector operations require equal lengths or a scalar
#>   operand; R-style recycling is not supported
```

The two recycling forms that quickr *does* support — scalars, and a
vector spanning the rows of a matrix — are exactly the ones with an
exact, non-partial interpretation:

```{r}
f <- quick(function(x, m) {
  declare(type(x = double(3)), type(m = double(3, 2)))
  x * m
})
f(c(1, 10, 100), matrix(as.double(1:6), 3, 2))
#>      [,1] [,2]
#> [1,]    1    4
#> [2,]   20   50
#> [3,]  300  600
```

### Matrix multiplication and linear algebra

`%*%`, `crossprod()`, `tcrossprod()`, `solve()`, and the triangular
solvers check conformability like R does. Dimensions known at compile
time are checked at compile time; unknown dimensions are checked at run
time before the underlying BLAS/LAPACK routine is called:

```{r}
f <- quick(function(a, b) {
  declare(type(a = double(n, k)), type(b = double(m, j)))
  a %*% b
})
f(matrix(1, 2, 3), matrix(1, 2, 2))
#> Error: non-conformable arguments in %*%
```

`solve(a, b)` requires a square `a`, as in R: a statically rectangular
system is a compile error, and when squareness depends on runtime sizes
it is checked before the solve. Use `qr.solve()` for least-squares
solutions of rectangular systems. The right-hand side must be a vector or
matrix. R also accepts a scalar right-hand side for a `1x1` system; quickr
refuses that rank-0 form. Declare a length-one vector when that case is
required.

`cbind()` and `rbind()` support scalar recycling and otherwise require the
common nonscalar extent to be provably equal at compile time. Arguments
declared with the same symbolic extent satisfy that requirement. Distinct or
otherwise unresolved extents are a compile error, even if their runtime values
would be equal.

## Differences from R

This section lists quickr's error divergences and known non-error
differences. If you find another case where a compiled function silently
returns something different from R, that is a bug — please report it.

### Integer and logical `prod()` results retain their type {#prod}

R returns a double from `prod()` when every argument is integer or
logical. quickr currently returns an integer in those cases, so both the
result type and integer-overflow behavior can differ. Use double inputs,
for example `prod(as.double(x))`, when R's double result is required.

### Integer `sum()` can overflow {#integer-sum-overflow}

R promotes an integer `sum()` result to double when the total exceeds the
integer range. quickr accumulates and returns an integer, so the result can
overflow instead. Convert the input to double before summing when the total
may exceed the integer range.

### Integer arithmetic can overflow

R detects overflow in integer `+`, `-`, and `*`, warns, and returns `NA`.
quickr does not support `NA` and uses Fortran integer arithmetic directly,
so an overflowing result can differ. Convert an operand to double when the
calculation can exceed the integer range.

### Zero divisors in `%%` and `%/%`

R returns `NA` or `NaN` with a warning for a zero divisor. Fortran's
remainder and division operations do not define it, so quickr refuses a
literal zero divisor at compile time and checks computed divisors at run time.

### Empty `min()` and `max()` inputs are rejected

R returns `Inf` from `min()` and `-Inf` from `max()` when the input is empty.
quickr does not support those identity values, so it rejects `min()` and
`max()` calls when every input is empty. Empty operands in a multi-argument
call are ignored when at least one other operand contributes a value.

Empty `which.min()` and `which.max()` inputs also differ. R returns
`integer(0)`; quickr returns the scalar index produced by Fortran's
`minloc()` or `maxloc()`. Do not pass empty inputs to these reductions.

### Whole-valued real exponents can differ for negative bases {#real-exponents}

R recognizes a whole-valued double exponent such as `3.0` when raising a
negative base. quickr emits a real-to-real power for a double exponent,
which can produce `NaN` instead of R's value. Use an integer exponent when
the exponent is integral and the base can be negative.

### No `NA`

`NA` values are not supported, in inputs or as literals. quickr does not
implement R's missing-value semantics, and compiled code does not detect
`NA` in inputs — passing `NA` produces undefined results. (For doubles,
`NaN` and `Inf` flow through arithmetic per IEEE rules as they do in R,
but no guarantees are made for code that branches on them.)

### No partial recycling

As described above: operand lengths must match exactly or one operand
must be a scalar (or a vector spanning a matrix's rows). R's general
recycling rule — including the case where one length divides the other,
which R accepts without a warning — is a compile-time error or a runtime
error instead. This also covers a `1x1` matrix in arithmetic with a
vector whose length is only known at run time: the compiled function
requires length 1 (returning a `1x1` matrix, as R does) and errors on a
longer vector, where R would recycle with a deprecation warning.

### Zero-length operands do not recycle to zero-length results

R arithmetic with a zero-length operand returns a zero-length result
regardless of the other operand's length: `numeric(0) + 1:3` is
`numeric(0)`. In quickr, a zero length is a length like any other. A
zero-length non-scalar operand is rejected, including when the other
operand is a scalar or another zero-length vector. This is a compile
error when the zero extent is known and a runtime error when it is not.

### Some zero-sized linear-algebra outputs are not supported

`%*%`, `crossprod()`, and `tcrossprod()` raise an error when
the result would have a zero extent, where R returns a zero-sized array.
A zero contracted dimension remains supported when every output extent is
nonzero, such as a `2x0` matrix multiplied by a `0x3` matrix; that result is
a `2x3` matrix of zeros, as in R. `outer()` and `%o%` reject zero-sized
outputs.

### `drop()` requires a fixed result rank

The rank of a compiled result cannot depend on a runtime extent. When a
matrix axis is not known at compile time, `drop()` accepts inputs whose
runtime extent is not 1 and raises an error when it is 1. Declare singleton
axes explicitly when `drop()` should remove them.

### A variable's type cannot change

`x <- x + 0.5` on an integer `x` is a compile-time error, as described
under [Types](#reassignment-cannot-narrow-a-type). R instead promotes
the variable to double.

### Complex values are elementwise-only {#complex}

Elementwise arithmetic and equality comparisons handle complex values
natively, including mixed integer, double, and complex operands:
`complex + double` promotes to complex, as in R. Everything else refuses:

- **Logical-complex pairs.** Elementwise operations do not promote a
  logical operand to complex. A logical-complex pair is a compile error
  where R returns a complex arithmetic result or a logical comparison.
- **Type joins.** `c()`, `ifelse()`, and `cbind()`/`rbind()` do not mix
  complex with other types — `c(x, y)` with a complex `x` and a double
  `y` is a compile error where R returns a complex vector.
- **Coercion.** `as.double()` refuses complex input instead of discarding
  the imaginary parts with a warning as R does.
- **Linear algebra.** `%*%`, `crossprod()`, `solve()`, `chol()`, and the
  rest of the linear-algebra surface refuse complex operands at compile
  time; quickr's lowerings are double-only, where R computes complex
  results.

(`t(x)` and `diag(x)` are mode-preserving and keep working on complex
values; order comparisons and `%%` on complex values are errors in R
itself, so refusing them is not a divergence.)

### `runif()` bound differences

R returns `NaN` values with a warning when `runif()` is called with
`min > max`. quickr applies its uniform scaling formula without checking
the bound order and returns values between the two bounds. Check the bounds
before calling `runif()` when their order is not known statically.

When `min` and `max` are equal, R returns the bound without consuming an RNG
value. quickr still draws and scales a uniform value, so the result is equal
but the subsequent RNG state differs.

### Scalar transpose {#scalar-transpose}

R's `t(x)` gives a scalar input `1x1` matrix dimensions. quickr returns the
scalar without dimensions. Vector and matrix transposes retain their
documented shapes.

### Subscript bounds are only partially checked

R validates every subscript: `x[0]` drops elements, negative subscripts
exclude, and out-of-range reads pad with `NA`. None of those have a
Fortran equivalent, and per-element bounds checks inside hot loops would
defeat the purpose of compiling, so quickr's contract is:

- Subscripts that are provably out of range at compile time (literal
  values, ranges provably outside a known extent) are compile errors.
  Zero and negative literal subscripts are rejected. Read and write
  subscripts (`x[i]` and `x[i] <- v`) get the same validation.
- A subscript whose value is only known at run time — a scalar `x[i]`,
  or a range `x[a:b]` with symbolic bounds — is **not** checked,
  matching Fortran's own contract: reading or writing out of bounds is
  undefined behavior, as in C. It will not return `NA`-padded results
  the way R does.

If you need R's bounds behavior, validate indices before the loop.

### `ifelse()` {#ifelse}

Four differences:

- **Test type.** R coerces numeric `test` values to logical. quickr requires
  `test` to already be logical and refuses other types.
- **Result type.** In R, `ifelse()`'s result type can depend on which
  branches are actually reached at run time:
  `ifelse(c(TRUE, TRUE), 1:2, c(0.5, 0.5))` is integer, because the `no`
  branch never materializes. quickr always promotes to the join of both
  branch types (double here), regardless of `test`'s values.
- **Result shape.** The result takes the shape of `test`, as in R. A
  scalar `test` with array-valued branches (where R would return a
  length-1 result sliced out of the array) is a compile error.
- **Branch lengths.** Branches shorter than `test` are recycled by R;
  quickr requires each branch to be a scalar or match `test`'s length
  (checked at run time when lengths are unknown). A branch known to be
  zero-length is a compile error.

```{r}
quick(function(t, a, b) {
  declare(type(t = logical(1)), type(a = double(3)), type(b = double(3)))
  ifelse(t, a, b)
})
#> Error: ifelse() result takes the shape of `test`; array-valued yes/no
#>   with scalar test is not supported
```

### Logical operators require logical operands {#logical-operators}

R coerces numeric operands to logical (`1 & 2` is `TRUE` and `!1` is
`FALSE`); quickr requires the operands of `!`, `&`, `|`, `&&`, and `||`
to already be logical and errors otherwise.

`&&` and `||` otherwise follow R's scalar semantics: operands must be
length 1 (a compile error when the length is known, or a runtime error
when it is not), and they short-circuit — the right operand is evaluated only
when the left side does not decide the answer, so a guarded access like
`i <= n && x[i] > 0` never evaluates `x[i]` out of range. For
elementwise logic over vectors, use `&` and `|`.

quickr still translates complete lazy branches. Recognized static elementwise
shape mismatches in a `&&` or `||` right operand, and wrong-arity calls to
supported built-ins in those operands or an `ifelse()` branch, are deferred with
the branch. A skipped path does not error; reaching it raises a runtime error.
Other unsupported expressions may still be rejected when `quick()` translates
the function.

### Local closures accept only pure argument expressions

R evaluates function arguments lazily. Local closures inside compiled
functions (including functions passed to `sapply()`) accept only argument
expressions that quickr can prove are pure, and evaluate accepted arguments
eagerly. Calls with effectful expressions such as `pair(runif(1), 2)` are
compile errors. Code that relies on an argument never being evaluated will
behave differently.

### Deprecation warnings are not reproduced

Where R accepts an operation but warns — recycling a `1x1` matrix in
arithmetic is the one such case in the supported subset — quickr
performs the operation without a warning.
