---
title: "quickr semantics: types, shapes, and differences from R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{quickr semantics: types, shapes, and differences from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

quickr compiles a subset of R. This vignette documents the semantics of
that subset: what type a result has, when shapes are checked, and every
place where a compiled function deliberately behaves differently from R.
All differences take the same form — an error where R would return a
value — never a different value.

## The contract

> `quick(f)(x)` either returns exactly what `f(x)` returns — same values,
> same `typeof()`, same shape — or raises an error. Never a third thing.

When quickr cannot reproduce R's behavior, it refuses. It prefers to
refuse at compile time, when `quick()` is called; when the decision
depends on values only known at run time (for example, sizes declared as
`NA` or as symbolic constraints like `double(n)`), the compiled function
performs a cheap runtime check and raises an error on mismatch. The text
of these error messages differs from R's.

The rest of this vignette is that contract in detail: first result
*types*, then operand *shapes*, then the list of deliberate divergences.

## Types

Value types follow R's promotion order for the supported atomic modes:

```
logical < integer < double < complex
```

Character values are not supported (declaring a `character` argument is
a compile error). Complex values support elementwise arithmetic and
equality comparisons; order comparisons (`<`, `<=`, `>`, `>=`) and `%%`
are errors, exactly as they are in R. Outside elementwise operations,
complex support is narrower than the lattice suggests — see
[complex values](#complex) under the differences below.

An operation's result type is the "join" of its operands' types — the
highest type among them — except where R fixes the result type, which
quickr then matches:

| Operation | Result type |
|---|---|
| `+` `-` `*` (and unary `+` `-`) | join of the operand types; logicals count as integers (in R, `TRUE + TRUE` is `2L`) |
| `/` | double (complex when an operand is complex, as in R) |
| `^` | double (complex when an operand is complex, as in R; see note below) |
| `%%`, `%/%` | join of the operand types; logicals count as integers |
| `<` `<=` `>` `>=` `==` `!=` | logical |
| `&`, `|`, `!`, `&&`, `||` | logical (see [logical operators](#logical-operators) below) |
| `c()`, and multi-argument `min()` / `max()` / `sum()` / `prod()` | join of all argument types |
| single-argument reductions (`sum(x)`, ...), `abs()` | type of `x`; logical counts as integer |
| `ifelse(test, yes, no)` | join of `yes` and `no` (but see [ifelse](#ifelse) below) |
| `%*%`, `crossprod()`, `solve()`, and other linear algebra | always double, as in R; complex operands are a compile error (see [complex values](#complex)) |
| `t(x)`, `diag(x)` | same type as `x`; the identity forms `diag(n)` are double, as in R |
| `cbind()` / `rbind()` | join of all argument types |

For example, mixing integer and double promotes:

```{r}
f <- quick(function(x, y) {
  declare(type(x = integer(1)), type(y = integer(1)))
  x / y
})
f(5L, 2L)
#> [1] 2.5
```

and logical arithmetic produces integers, as in R:

```{r}
f <- quick(function(x, y) {
  declare(type(x = logical(1)), type(y = logical(1)))
  x + y
})
r <- f(TRUE, TRUE)
r
#> [1] 2
typeof(r)
#> [1] "integer"
```

**A note on `^`:** the result is always double, as in R, but when the
exponent is a whole number quickr computes the power with an integer
exponent. This is exact, and well-defined for negative bases:
`(-2) ^ 3L` is exactly `-8`.

### Reassignment cannot narrow a type

R happily re-binds a name to a value of a wider type: after
`x <- x + 0.5`, an integer `x` becomes double. Compiled code cannot
re-type a variable, and silently truncating the value would change
results, so quickr makes this a compile-time error:

```{r}
quick(function(x) {
  declare(type(x = integer(1)))
  x <- x + 0.5
  x
})
#> Error: cannot reassign `x`: assignment would narrow double to integer;
#>   R would promote `x` to double
```

The same check applies to subassignment (`x[i] <- ...`) and
superassignment (`x <<- ...`, `x[i] <<- ...`). Widening in the other
direction (assigning an integer value to a double variable) is fine and
inserts the conversion R would perform.

## Shapes

Every operation that combines two or more values checks that their
shapes are compatible. A shape question has one of three answers, and
each answer has a fixed consequence:

- **Provably compatible** at compile time — allowed, no check emitted.
- **Provably incompatible** at compile time — compile error.
- **Unknown** at compile time (sizes declared `NA`, or symbolic sizes
  that cannot be proven equal) — the compiled function checks at run
  time and raises an error on mismatch.

There is no case where a mismatch is silently accepted.

For elementwise operations (arithmetic, comparisons, `&`, `|`),
"compatible" means:

| Operands | Behavior |
|---|---|
| scalar with anything | allowed; the scalar is broadcast |
| identical shapes | allowed |
| vector with matrix, `length(vec) == nrow(mat)` | allowed; the vector is recycled column-wise, exactly as in R |
| a `1x1` matrix in *arithmetic* | treated as a scalar (R allows this, with a deprecation warning) |
| a `1x1` matrix in comparisons and `&` / `|` | treated as a one-row matrix; mismatched shapes are rejected, and R errors here too |
| anything else — unequal lengths, even when one divides the other | **error**: R-style recycling is not supported |

The last row is the important one. In R, `1:6 + c(10, 20)` recycles the
shorter vector. quickr never does this:

```{r}
quick(function(x, y) {
  declare(type(x = double(4)), type(y = double(2)))
  x + y
})
#> Error: elementwise vector operations require equal lengths or a scalar
#>   operand; R-style recycling is not supported
```

When lengths are not known at compile time, the same rule is enforced at
run time:

```{r}
f <- quick(function(x, y) {
  declare(type(x = double(n)), type(y = double(m)))
  x + y
})
f(c(1, 2), c(3, 4))
#> [1] 4 6
f(c(1, 2, 3), c(1, 2))
#> Error: elementwise vector operations require equal lengths or a scalar
#>   operand; R-style recycling is not supported
```

The two recycling forms that quickr *does* support — scalars, and a
vector spanning the rows of a matrix — are exactly the ones with an
exact, non-partial interpretation:

```{r}
f <- quick(function(x, m) {
  declare(type(x = double(3)), type(m = double(3, 2)))
  x * m
})
f(c(1, 10, 100), matrix(as.double(1:6), 3, 2))
#>      [,1] [,2]
#> [1,]    1    4
#> [2,]   20   50
#> [3,]  300  600
```

### Matrix multiplication and linear algebra

`%*%`, `crossprod()`, `tcrossprod()`, `solve()`, and the triangular
solvers check conformability like R does. Dimensions known at compile
time are checked at compile time; unknown dimensions are checked at run
time before the underlying BLAS/LAPACK routine is called:

```{r}
f <- quick(function(a, b) {
  declare(type(a = double(n, k)), type(b = double(m, j)))
  a %*% b
})
f(matrix(1, 2, 3), matrix(1, 2, 2))
#> Error: non-conformable arguments in %*%
```

`solve(a, b)` requires a square `a`, as in R: a statically rectangular
system is a compile error, and when squareness depends on runtime sizes
it is checked before the solve. Use `qr.solve()` for least-squares
solutions of rectangular systems.

## Differences from R

Everything listed here is an *error* divergence: a compiled function
refuses where plain R would produce a value. If you find a case where a
compiled function silently returns something different from R, that is a
bug — please report it.

### No `NA`

`NA` values are not supported, in inputs or as literals. quickr does not
implement R's missing-value semantics, and compiled code does not detect
`NA` in inputs — passing `NA` produces undefined results. (For doubles,
`NaN` and `Inf` flow through arithmetic per IEEE rules as they do in R,
but no guarantees are made for code that branches on them.)

### No partial recycling

As described above: operand lengths must match exactly or one operand
must be a scalar (or a vector spanning a matrix's rows). R's general
recycling rule — including the case where one length divides the other,
which R accepts without a warning — is a compile-time error or a runtime
error instead.

### Zero-length operands do not recycle to zero-length results

R arithmetic with a zero-length operand returns a zero-length result
regardless of the other operand's length: `numeric(0) + 1:3` is
`numeric(0)`. In quickr, a zero length is a length like any other. An
operand *known* at compile time to be zero-length combines only with a
scalar (`double(0) + 1` is `numeric(0)`, matching R); combining it with
any other vector — even another known-zero-length one — is a compile
error. When lengths are only known at run time, a zero-length operand
errors unless the other operand's length is also zero (two zero-length
operands produce a zero-length result, as in R).

### A variable's type cannot change

`x <- x + 0.5` on an integer `x` is a compile-time error, as described
under [Types](#reassignment-cannot-narrow-a-type). R instead promotes
the variable to double.

### Complex values are elementwise-only {#complex}

Elementwise arithmetic and equality comparisons handle complex values
natively, including mixed operands: `complex + double` promotes to
complex, as in R. Everything else refuses:

- **Type joins.** `c()`, `ifelse()`, and `cbind()`/`rbind()` do not mix
  complex with other types — `c(x, y)` with a complex `x` and a double
  `y` is a compile error where R returns a complex vector.
- **Linear algebra.** `%*%`, `crossprod()`, `solve()`, `chol()`, and the
  rest of the linear-algebra surface refuse complex operands at compile
  time; quickr's lowerings are double-only, where R computes complex
  results.

(`t(x)` and `diag(x)` are mode-preserving and keep working on complex
values; order comparisons and `%%` on complex values are errors in R
itself, so refusing them is not a divergence.)

### Subscript bounds are only partially checked

R validates every subscript: `x[0]` drops elements, negative subscripts
exclude, and out-of-range reads pad with `NA`. None of those have a
Fortran equivalent, and per-element bounds checks inside hot loops would
defeat the purpose of compiling, so quickr's contract is:

- Subscripts that are provably out of range at compile time (literal
  values, ranges provably outside a known extent) are compile errors.
  Zero and negative literal subscripts are rejected. Read and write
  subscripts (`x[i]` and `x[i] <- v`) get the same validation.
- A subscript whose value is only known at run time — a scalar `x[i]`,
  or a range `x[a:b]` with symbolic bounds — is **not** checked,
  matching Fortran's own contract: reading or writing out of bounds is
  undefined behavior, as in C. It will not return `NA`-padded results
  the way R does.

If you need R's bounds behavior, validate indices before the loop.

### `ifelse()` {#ifelse}

Three differences:

- **Result type.** In R, `ifelse()`'s result type can depend on which
  branches are actually reached at run time:
  `ifelse(c(TRUE, TRUE), 1:2, c(0.5, 0.5))` is integer, because the `no`
  branch never materializes. quickr always promotes to the join of both
  branch types (double here), regardless of `test`'s values.
- **Result shape.** The result takes the shape of `test`, as in R. A
  scalar `test` with array-valued branches (where R would return a
  length-1 result sliced out of the array) is a compile error.
- **Branch lengths.** Branches shorter than `test` are recycled by R;
  quickr requires each branch to be a scalar or match `test`'s length
  (checked at run time when lengths are unknown). A branch known to be
  zero-length is a compile error.

```{r}
quick(function(t, a, b) {
  declare(type(t = logical(1)), type(a = double(3)), type(b = double(3)))
  ifelse(t, a, b)
})
#> Error: ifelse() result takes the shape of `test`; array-valued yes/no
#>   with scalar test is not supported
```

### Logical operators require logical operands {#logical-operators}

R coerces numeric operands to logical (`1 & 2` is `TRUE`); quickr
requires the operands of `&`, `|`, `&&`, and `||` to already be logical
and errors otherwise.

`&&` and `||` otherwise follow R's scalar semantics: operands must be
length 1 (a compile error in quickr; R raises the equivalent error at
run time), and they short-circuit — the right operand is evaluated only
when the left side does not decide the answer, so a guarded access like
`i <= n && x[i] > 0` never evaluates `x[i]` out of range. For
elementwise logic over vectors, use `&` and `|`.

### Closure arguments are evaluated eagerly

R evaluates function arguments lazily. Local closures inside compiled
functions (including functions passed to `sapply()`) evaluate their
arguments eagerly. Code that relies on an argument never being evaluated
will behave differently.

### Deprecation warnings are not reproduced

Where R accepts an operation but warns — recycling a `1x1` matrix in
arithmetic is the one such case in the supported subset — quickr
performs the operation without a warning.
